On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification

### 2020-07-02 paper/cav_determinisim_main.tex
revision of first page and addition of CyRes to text and bib

% ***************************************************
%  TITLE
% ***************************************************
\title{Investigating Determinism of Game Engines for Autonomous Vehicle Verification}
\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
\author[1,3]{Greg Chance}
\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
\author[2,3]{Tony Pipe}
\author[1,3]{Kerstin Eder}
\affil[1]{Trustworthy Systems Lab, University of Bristol, Bristol, UK}
\affil[2]{University of West of England, Bristol, UK}
\affil[3]{Bristol Robotics Lab, Bristol, UK}
\maketitle

\begin{abstract}
\noindent 
The autonomous vehicle (AV) community is adopting the use of game engines to develop vehicle control systems and testing environments using simulation. Determinism is a key requirement for simulation during AV development and testing. In particular, it ensures repeatability of tests, so that software bugs can be found and fixed, and simulation results are trustworthy.
%
% KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
% These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
%
However, game engines do not need to be deterministic per se. % when used for games. 
%
This work identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation variance} for specific configurations and utilisation of the software and hardware, respectively. 
%
A case study using Carla, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from a small scenario indicate the presence of deterministic operational domains. \todo[inline]{KIE: review at end}

@@

\begin{IEEEkeywords}
Autonomous Driving, Determinism, Physics Engines, Autonomous Vehicles, Verification and Validation (V\&V), Simulation, Testing
\end{IEEEkeywords}
\IEEEpeerreviewmaketitle

% ***************************************************
%  INTRODUCTION
% ***************************************************
\section{Introduction} \label{s:introduction}
% What is verification
\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it. These simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving assertions that must not be violated.

@@

% Requirements of simulation based verification SBV
The benefit of using state-of-the-art game engines as simulators is that they offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses. 
%
Furthermore, they are easy to setup and run; they also are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}. 
%
\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}
%
For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models. Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development. 
\todo[inline]{GC: what is ray-case?}
%
Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013} may also be required.

% Game engines for SBV
Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation. 
%
\todo[inline]{GC Replace: However, several challenges remain; one of these is determinism. with: However, game engines are designed principally for user experience whereas the requirements for CAV verification is not user experience but determinism.}
%
% What is a deterministic system
Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
%
Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario.

@@

% Verification and Determinism 
% Why is determinism important for verification
%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
Determinism is a key prerequisite for simulation during AV development and testing. 
%
A deterministic simulation environment guarantees that tests are repeatable. 
%
Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy. 
%
%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution. Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour. 
%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
%
If the simulation is not deterministic, however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs. In the worst case, bugs that could have been identified in simulation are left undetected, resulting in false confidence in the safety of the AV's control software.

>>>>

% ***************************************************
%  TITLE
% ***************************************************
\title{On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification}
\author[1,3]{Abanoub Ghobrial\thanks{$^{1}${\footnotesize \{abanoub.ghobrial, greg.chance, kerstin.eder\}@bristol.ac.uk}}}
\author[1,3]{Greg Chance}
\author[2,3]{Severin Lemaignan\thanks{$^{2}${\footnotesize \{severin.lemaignan, tony.pipe\}@brl.ac.uk}}}
\author[2,3]{Tony Pipe}
\author[1,3]{Kerstin Eder}
\affil[1]{Trustworthy Systems Laboratory, University of Bristol, Bristol, UK}
\affil[2]{University of West of England, Bristol, UK}
\affil[3]{Bristol Robotics Laboratory, Bristol, UK}
\maketitle

\begin{abstract}
\noindent 
Game engines are increasingly used as simulation platforms by the autonomous vehicle (AV) community to develop vehicle control systems and test environments. 
%
Determinism is a key requirement for simulation-based development and verification; it ensures tests are repeatable, so that software bugs can be found, fixed and fixes can be checked, and simulation results are trustworthy.
%
% KIE: Ideally, these engines are deterministic, but we've found that they are not. :(
% These engines must be deterministic for finding and resolving defects (bugs) in software and ensuring auditable and repeatable tests can be executed for verification and validation (V\&V). 
%
However, game engines do not need to be deterministic per se. % when used for games. 
%
This paper identifies the requirements these engines need to fulfil when used for AV development and testing, and proposes a method to determine the \textit{simulation variance} for specific configurations and utilisation of the software and hardware, respectively. 
%
A case study using Carla, an open-source simulation environment powered by Unreal Engine, illustrates the proposed method and results from an example scenario indicate the presence of deterministic operational domains. \todo{Review again at end. Potentially give away more on results to motivate the reader to continue.}

@@

\begin{IEEEkeywords}
Autonomous Driving, Autonomous Vehicles, Determinism, Physics Engines,  Verification and Validation (V\&V), Simulation, Testing
\end{IEEEkeywords}
\IEEEpeerreviewmaketitle

% ***************************************************
%  INTRODUCTION
% ***************************************************
\section{Introduction} \label{s:introduction}
% What is verification
% Can't use this again as this is how the AI Test paper starts. -> Avoid self-plagiarism!
%\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. 
%
\IEEEPARstart{S}{}imulation-based techniques can be used to verify autonomous driving functions under development benefiting from full control over the road network and the actors within it. Simulated tests aim to provide evidence to developers and regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988} which form a body of driving rules, termed assertions, that must not be violated.

@@

% Requirements of simulation based verification SBV
State-of-the-art game engines provide a convenient option for simulation-based testing. They offer sufficient realism~\cite{Koopman2018} in the physical domain combined with realistic rendering of scenes, potentially suitable for perception stack testing and visual inspection of accidents or near misses. 
%
Furthermore, they are easy to setup and run; they also are simple to control and observe, both with respect to the environment the AV operates in as well as the temporal development of actors~\cite{Ulbrich2015}. 
%
%\todo[inline]{These game engines provide a convenient route to testing ego control and perception stack in a realistic environment}

For game engines to be used in simulation-based testing of AVs, they must offer sensors that operate at an accuracy equivalent to those on the AV, e.g.\ video cameras, LIDAR, IR, etc., or interfaces for developers to integrate their own sensor models. Carla currently offers various basic types of cameras and ray-case based lidar; extensions are still under development. 
\todo{GC: what is ray-case? KIE: Not sure. :(}
%
Finally, support for hardware-in-the-loop development or a real-time test-bed for cyber-security testing~\cite{Javaid2013,cyres} may also be required.

% Game engines for SBV
Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation. 
%
\todo{GC Replace: However, several challenges remain; one of these is determinism. with:}
%
However, while game engines are designed primarily for user experience, the requirements for AV verification go beyond that and include determinism.
%
% What is a deterministic system
In this context, deterministic refers to the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}.
%
Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario.

@@

% Verification and Determinism 
% Why is determinism important for verification
%For game engines to be useful as a verification tool then they must be deterministic. -> any sim tool must be deterministic, not just game engines ;)
\todo{move to earlier :)}Determinism is a key prerequisite for simulation during AV development and testing. 
%
A deterministic simulation environment guarantees that tests are repeatable. 
%
Thus, coverage results are stable and, when a test fails, debugging can rely on the test producing the same trace and outcome when repeated. This ensures that software bugs can be found and fixed efficiently, and that simulation results are trustworthy. 
%
%Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution. Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe behaviour. 
%When considering the development of autonomous vehicles, untested code could lead to dangerous behaviours not being identified and fatal road accidents may occur.
%
If the simulation is not deterministic, however, variance in actor positions, for instance, may lead to assessment errors, making it difficult to understand and remove bugs. In the worst case, bugs that could have been identified in simulation are left undetected, resulting in false confidence in the safety of the AV's control software.
